---
title: "Analysis Report Regarding Questions of COVID-19"
author: "William"
date: "2/22/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
#source("analysis.R")
library("knitr")
```

## Section 1. Problem Domain Description
COVID-19, which is short for coronavirus disease of 2019, is the illness caused by the SARS-CoV-2 virus first identified in Wuhan, China in December of 2019. Since then, the virus has rapidly spread across the world, leading the World Health Organization to declare a global pandemic. Millions of Americans have been infected by the virus, and hundreds of thousands have died due to the disease with those numbers only continuing to grow each day. A global race to develop a vaccine in record-breaking time ensued, with over 100 different candidates being tested across the globe. Despite multiple vaccines receiving emergency authorizations from multiple different nations, the situation is worsening daily as new mutant strains are being identified such as those identified in the United Kingdom. In the United States, public health officials are struggling to convince the populous that the vaccines are safe and effective, leading to widespread anti-vaccine protests seeking to slow the vaccination efforts, which only lends itself to give the virus more time to develop a mutation to defeat the current vaccine formulations.

Thus, analyzing data related to COVID-19 is worthwhile since it will help people understand the overall situation and severity of the pandemic and arouse their interest in adopting protective measures like mask-wearing, social-distancing, and vaccination. In addition, analyzing this data may expose differences in the ability of different regulations between states to contain the virus, which may prove beneficial in helping state governments are only utilizing restrictions that truly work to contain this pathogen.

## Section 2. Data Description
```{r section_2, include = FALSE}
source("section_2.R")
```

#### NYT Mask-Wearing Survey
The [NYT Mask-Wearing Survey data set](https://github.com/nytimes/covid-19-data/tree/master/mask-use) contains estimates of mask-usage from 250,000 survey responses for each county in the US. Each participant was asked "How often do you wear a mask in public when you expect to be within six feet of another person?" and given the choices of never, rarely, sometimes, frequently, or always. The survey was done in 2020 from July 2 to July 14, and was assembled by The New York Times and Dynata.

```{r nyt_mask_wearing_survey_sample}
kable(head(mask_use_by_county_df))
```

The `COUNTYFP` column is the FIPS code for the county, and the rest of the columns are estimates for the percent of people in that county who responded with that option. Those values always add up to about one.

#### CDC COVID-19 Vaccinations in the United States
The [COVID-19 Vaccinations in the United States data set](https://covid.cdc.gov/covid-data-tracker/#vaccinations) contains number of vaccine doses administered by state. Data on COVID-19 vaccine doses administered in the United States are collected by vaccination providers and reported to CDC through multiple sources, including jurisdictions, pharmacies, and federal entities, which use various reporting methods, including Immunization Information Systems, Vaccine Administration Management System, and direct data submission. 

```{r vaccination_by_state_sample}
kable(head(vaccination_by_state_df))
```

- `Total doses administered` column is the total number of vaccine doses that have been given to people. 

- `Doses administered per 100k` column is the total number of vaccine doses given for every 100,000 people.

- `18+ Doses Administered` column is the total number of vaccine doses that have been given to people for the overall population

- `18+ Doses administered per 100k` column is the total number of vaccine doses given for every 100,000 people aged 18 years and older.


### Subsection 2.2 Summary Analysis

#### NYT Mask-Wearing Survey
Grouped by counties, an average of `r round(mean(mask_use_by_county_df$ALWAYS) * 100)`% of the responses are "Always," and an average of `r round(mean(mask_use_by_county_df$NEVER) * 100)`% of the responses are "Never." For a single county, the values for each response are supposed to sum to one. In reality, the values are rounded to three decimal places, so the sum for each county ranges from `r min(mask_use_by_county_sum_df$sum)` to `r max(mask_use_by_county_sum_df$sum)`.

```{r mask_use_by_county_summary_df}
kable(mask_use_by_county_summary_df)
```

```{r mask_use_violin_plot}
mask_use_violin_plot
```

```{r mask_use_always_map}
mask_use_always_map
```

```{r mask_use_never_map}
mask_use_never_map
```

There doesn't seem to be any significant outliers. This is probably because there were 250,000 survey responses for a survey with only 5 options. Any individual county would have to have a lot of different responses in order to be able to become an outlier. Also, there is less chance for outliers because this data set was grouped into counties, forcing all of the columns for each row to sum to one. There are no NA values, and it seems to have data for almost every county.

#### CDC COVID-19 Vaccinations in the United States
Grouped by states, there are an average of `r mean(vaccination_by_state_df$Doses_Administerd_per_100k, na.rm = T)` per 100,000 (`r mean(vaccination_by_state_df$Doses_Administerd_per_100k, na.rm = T) * 100/100000`%) of population in each state got doses given. The number of doses administered per 100,000 ranges from `r min(mask_use_by_county_sum_df$Doses_Administerd_per_100k)` to `r max(mask_use_by_county_sum_df$Doses_Administerd_per_100k)`.

```{r vaccination_by_state_summary_df}
kable(vaccination_by_state_summary_df)
```

## Section 3: Specific Question Analysis

